---
title: "Chapter 3: Kmeans"
output: html_document
---
```{r libs, message=FALSE, warning=FALSE}
library(ggplot2)
library(RColorBrewer)
library(daltoolbox)
library(harbinger)
source("https://raw.githubusercontent.com/eogasawara/TSED/refs/heads/main/code/header.R")
```
## Theoretical Overview
This example focuses on k-means clustering. Clustering time windows and labeling outliers can identify unusual behavior segments.
## Example Overview and Goals
We will: load a labeled toy series, configure k-means on windowed features, detect cluster outliers, evaluate, and visualize.
### What You Will Do
Prepare the environment, run k-means-based detection, print/evaluate detections, and plot.
### Setup and Libraries
Load helpers and packages.
```{r setup, message=FALSE, warning=FALSE}
```
### Data Loading and Prep
Load a toy anomalies dataset.
```{r data_loading}
data("examples_anomalies")
dataset <- examples_anomalies$tt_warped
```
### Other Steps
Additional supporting steps that glue the workflow.
```{r model_fit}
plot_ts(x = seq_along(dataset$serie), y = dataset$serie)
model <- hanct_kmeans(1)
model <- fit(model, dataset$serie)
```
### Event Detection
Run the detector to obtain event flags or scores.
```{r detect}
detection <- detect(model, dataset$serie)
```
### Other Steps
Additional supporting steps that glue the workflow.
```{r inspect}
print(detection |> dplyr::filter(event == TRUE))
```
### Evaluation
Compute evaluation metrics or diagnostics if ground truth is available.
```{r evaluate}
evaluation <- daltoolbox::evaluate(model, detection$event, dataset$event)
```
### Other Steps
Additional supporting steps that glue the workflow.
```{r evaluate_print}
print(evaluation$confMatrix)
```
### Visualization and Output
Plot the series with detected events and optionally save figures. Combine ggplot layers in one chunk for clarity.
```{r viz_plot}
grf <- har_plot(model, dataset$serie, detection, dataset$event)
```
### Other Steps
Additional supporting steps that glue the workflow.
```{r viz_layers}
grf <- grf + font
```
### Visualization and Output
Plot the series with detected events and optionally save figures. Combine ggplot layers in one chunk for clarity.
```{r viz_save}
#save_png(grf, "figures/chap3_kmeans.png", 1280, 720)
grf
```
## References
* MacQueen, J. (1967). Some methods for classification and analysis of multivariate observations.
* Ogasawara, E., Salles, R., Porto, F., Pacitti, E. Event Detection in Time Series. Springer, 2025. doi:10.1007/978-3-031-75941-3.
