---
title: "Chapter 2: Emd Decomposition"
output: html_document
---
```{r libs, message=FALSE, warning=FALSE}
library(ggplot2)
library(RColorBrewer)
library(harbinger)
library(hht)
library(patchwork)
source("https://raw.githubusercontent.com/eogasawara/TSED/refs/heads/main/code/header.R")
```
## Theoretical Overview
This example focuses on Empirical Mode Decomposition (EMD). EMD decomposes signals into intrinsic mode functions, separating oscillatory modes for analysis and detection.
## Example Overview and Goals
We demonstrate a complete, reproducible workflow: setting up libraries, loading data, configuring a detector, fitting it, running detection, optionally evaluating, and visualizing the results.
### Other Steps
Additional supporting steps that glue the workflow.
```{r}
# Chapter 2: Emd Decomposition
# Overview:
# - Loads example dataset (if applicable)
# - Fits a detector and runs event detection
# - Plots results with events overlaid
#
```
### Setup and Libraries
Short rationale for the libraries and any project-specific sources.
```{r setup, message=FALSE, warning=FALSE}
```
### Other Steps
Additional supporting steps that glue the workflow.
```{r}
options(scipen = 999)
```
### Setup and Libraries
Short rationale for the libraries and any project-specific sources.
```{r}
```
### Other Steps
Additional supporting steps that glue the workflow.
```{r}
set.seed(100)
# Load example dataset
```
### Data Loading and Prep
Read the dataset and perform any minimal preparation required for modeling.
```{r data_loading}
data(examples_harbinger)
```
### Other Steps
Additional supporting steps that glue the workflow.
```{r}
data <- examples_harbinger$global_temperature_yearly
data$event <- FALSE
y <- data$serie
yts <- ts(y, start = c(1850, 1))
xts <- time(yts)
id <- 1:length(yts)
model <- hht::CEEMD(yts, id, verbose = FALSE, 0.1, 1)
imfs <- model[["imf"]]
```
### Visualization and Output
Plot the series with detected events and optionally save figures. Combine ggplot layers in one chunk for clarity.
```{r}
grf <- autoplot(yts)
```
### Other Steps
Additional supporting steps that glue the workflow.
```{r}
grf <- grf + theme_bw(base_size = 10)
grf <- grf + theme(plot.title = element_blank())
grf <- grf + theme(panel.grid.major = element_blank()) + theme(panel.grid.minor = element_blank())
grf <- grf + ylab("temperature")
grf <- grf + xlab("time")
grf <- grf + theme(plot.caption = element_text(hjust = 0.5))
grf <- grf  + font
grfts <- grf
```
### Visualization and Output
Plot the series with detected events and optionally save figures. Combine ggplot layers in one chunk for clarity.
```{r}
grf <- autoplot(ts(imfs[,1], start = c(1850, 1)))
```
### Other Steps
Additional supporting steps that glue the workflow.
```{r}
grf <- grf + theme_bw(base_size = 10)
grf <- grf + theme(plot.title = element_blank())
grf <- grf + theme(panel.grid.major = element_blank()) + theme(panel.grid.minor = element_blank())
grf <- grf + ylab("imf 1")
grf <- grf + xlab("time")
grf <- grf + theme(plot.caption = element_text(hjust = 0.5))
grf <- grf  + font
grf1 <- grf
```
### Visualization and Output
Plot the series with detected events and optionally save figures. Combine ggplot layers in one chunk for clarity.
```{r}
grf <- autoplot(ts(imfs[,2], start = c(1850, 1)))
```
### Other Steps
Additional supporting steps that glue the workflow.
```{r}
grf <- grf + theme_bw(base_size = 10)
grf <- grf + theme(plot.title = element_blank())
grf <- grf + theme(panel.grid.major = element_blank()) + theme(panel.grid.minor = element_blank())
grf <- grf + ylab("imf 2")
grf <- grf + xlab("time")
grf <- grf + theme(plot.caption = element_text(hjust = 0.5))
grf <- grf  + font
grf2 <- grf
```
### Visualization and Output
Plot the series with detected events and optionally save figures. Combine ggplot layers in one chunk for clarity.
```{r}
grf <- autoplot(ts(imfs[,3], start = c(1850, 1)))
```
### Other Steps
Additional supporting steps that glue the workflow.
```{r}
grf <- grf + theme_bw(base_size = 10)
grf <- grf + theme(plot.title = element_blank())
grf <- grf + theme(panel.grid.major = element_blank()) + theme(panel.grid.minor = element_blank())
grf <- grf + ylab("imf 3")
grf <- grf + xlab("time")
grf <- grf + theme(plot.caption = element_text(hjust = 0.5))
grf <- grf  + font
grf3 <- grf
```
### Visualization and Output
Plot the series with detected events and optionally save figures. Combine ggplot layers in one chunk for clarity.
```{r}
grf <- autoplot(ts(imfs[,4], start = c(1850, 1)))
```
### Other Steps
Additional supporting steps that glue the workflow.
```{r}
grf <- grf + theme_bw(base_size = 10)
grf <- grf + theme(plot.title = element_blank())
grf <- grf + theme(panel.grid.major = element_blank()) + theme(panel.grid.minor = element_blank())
grf <- grf + ylab("imf 4")
grf <- grf + xlab("time")
grf <- grf + theme(plot.caption = element_text(hjust = 0.5))
grf <- grf  + font
grf4 <- grf
```
### Visualization and Output
Plot the series with detected events and optionally save figures. Combine ggplot layers in one chunk for clarity.
```{r}
grf <- autoplot(ts(imfs[,5], start = c(1850, 1)))
```
### Other Steps
Additional supporting steps that glue the workflow.
```{r}
grf <- grf + theme_bw(base_size = 10)
grf <- grf + theme(plot.title = element_blank())
grf <- grf + theme(panel.grid.major = element_blank()) + theme(panel.grid.minor = element_blank())
grf <- grf + ylab("imf 5")
grf <- grf + xlab("time")
grf <- grf + theme(plot.caption = element_text(hjust = 0.5))
grf <- grf  + font
grf5 <- grf
```
### Visualization and Output
Plot the series with detected events and optionally save figures. Combine ggplot layers in one chunk for clarity.
```{r}
grf <- autoplot(ts(imfs[,6], start = c(1850, 1)))
```
### Other Steps
Additional supporting steps that glue the workflow.
```{r}
grf <- grf + theme_bw(base_size = 10)
grf <- grf + theme(plot.title = element_blank())
grf <- grf + theme(panel.grid.major = element_blank()) + theme(panel.grid.minor = element_blank())
grf <- grf + ylab("imf 6")
grf <- grf + xlab("time")
grf <- grf + theme(plot.caption = element_text(hjust = 0.5))
grf <- grf  + font
grf6 <- grf
```
### Visualization and Output
Plot the series with detected events and optionally save figures. Combine ggplot layers in one chunk for clarity.
```{r}
grf <- autoplot(ts(model$residue, start = c(1850, 1)))
```
### Other Steps
Additional supporting steps that glue the workflow.
```{r}
grf <- grf + theme_bw(base_size = 10)
grf <- grf + theme(plot.title = element_blank())
grf <- grf + theme(panel.grid.major = element_blank()) + theme(panel.grid.minor = element_blank())
grf <- grf + ylab("residual")
grf <- grf + xlab("time")
grf <- grf + geom_point(size = 0.5, col="black") 
grf <- grf + theme(plot.caption = element_text(hjust = 0.5))
grf <- grf  + font
grfr <- grf
```
### Visualization and Output
Plot the series with detected events and optionally save figures. Combine ggplot layers in one chunk for clarity.
```{r}
grf <- wrap_plots(grfts, grf1, grf2, grf3, grf4, grf5, grf6, grfr, ncol = 3, widths = c(1, 1, 1), heights = c(1, 1, 1))
#save_png(grf, "figures/chap2_emd_decomposition.png", width = 1280, height = 1280)
grf
```
### Other Steps
Additional supporting steps that glue the workflow.
## References
* Huang, N. E., et al. (1998). The empirical mode decomposition and the Hilbert spectrum for nonlinear and non-stationary time series analysis.
* Ogasawara, E., Salles, R., Porto, F., Pacitti, E. Event Detection in Time Series. Springer, 2025. doi:10.1007/978-3-031-75941-3.
