---
title: "Chapter 2: Norm"
output: html_document
---
```{r libs, message=FALSE, warning=FALSE}
library(ggplot2)
library(RColorBrewer)
library(ggpmisc)
library(daltoolbox)
library(harbinger)
source("https://raw.githubusercontent.com/eogasawara/TSED/refs/heads/main/code/header.R")
```
## Theoretical Overview
This example focuses on normalization and scaling for time series modeling. We illustrate global min-max, differencing with scaling, sliding-window min-max, and anomaly normalization.
## Example Overview and Goals
We will: set up libraries, load data, prepare a sliding-window slice, apply several normalizations, and visualize their effects.
### What You Will Do
You will: prepare the environment, build a sliding window, apply different normalizations, and visualize them side by side.
### Setup and Libraries
Short rationale for the libraries and any project-specific sources.
```{r setup, message=FALSE, warning=FALSE}
options(scipen = 999)
```
### Other Steps
Additional supporting steps that glue the workflow.
```{r}
options(scipen=999)
```
### Setup and Libraries
Short rationale for the libraries and any project-specific sources.
```{r}
```
### Other Steps
Additional supporting steps that glue the workflow.
```{r}
```
### Setup and Libraries
Short rationale for the libraries and any project-specific sources.
```{r}
```
### Other Steps
Additional supporting steps that glue the workflow.
```{r}
# Load example dataset
```
### Data Loading and Prep
Read the dataset and perform any minimal preparation required for modeling.
```{r data_loading}
data(examples_harbinger)
```
### Other Steps
Additional supporting steps that glue the workflow.
```{r prep_windows}
data <- examples_harbinger$global_temperature_yearly
ats <- function(y) {
  yts <- ts(y, frequency=1, start = c(1850, 1))
  return(yts)  
}
y <- ats(data$serie)
x <- time(y)
sw_size <- 5
test_size <- 10 
range <- length(y)-77-1
swy <- ts_data(y, sw_size)
sw_range <- swy[(nrow(swy)-range+1),]
```
### Visualization and Output
Plot the series with detected events and optionally save figures. Combine ggplot layers in one chunk for clarity.
```{r viz_series}
grf <- autoplot(y) + theme_bw(base_size = 10) + geom_point(size=1)
```
### Other Steps
Additional supporting steps that glue the workflow.
```{r}
grf <- grf + theme(plot.title = element_blank())
grf <- grf + theme(panel.grid.major = element_blank()) + theme(panel.grid.minor = element_blank())
grf <- grf + ylab("temperature")
grf <- grf + xlab("year")
grf <- grf + geom_vline(xintercept = x[length(x)-sw_size+1-range+1], col="darkgreen", linewidth = 0.5, linetype="dotted")
grf <- grf + geom_vline(xintercept = x[length(x)-range+1], col="darkgreen", linewidth = 0.5, linetype="dotted")
grf <- grf + labs(caption = "(a)") 
grf <- grf + theme(plot.caption = element_text(hjust = 0.5))
grf <- grf  + font
grfS <- grf
### ts_gminmax
gen_graphics <- function(dgminmax, ddiff, dswminmax, d_an) {
  grf <- autoplot(ts(as.vector(dgminmax[nrow(dgminmax),]))) + theme_bw(base_size = 10) + geom_point(size=1)
  grf <- grf + theme(plot.title = element_blank())
  grf <- grf + theme(panel.grid.major = element_blank()) + theme(panel.grid.minor = element_blank())
  grf <- grf + ylab("")
  grf <- grf + xlab("global min-max")
  grf <- grf + theme(plot.caption = element_text(hjust = 0.5))
  grf <- grf  + font
  grf <- grf + ylim(0, 1)
  grfGA <- grf
  grf <- autoplot(ts(as.vector(ddiff[nrow(ddiff),]))) + theme_bw(base_size = 10) + geom_point(size=1)
  grf <- grf + theme(plot.title = element_blank())
  grf <- grf + theme(panel.grid.major = element_blank()) + theme(panel.grid.minor = element_blank())
  grf <- grf + ylab("")
  grf <- grf + xlab("diff with scaling")
  grf <- grf + theme(plot.caption = element_text(hjust = 0.5))
  grf <- grf  + font
  grf <- grf + ylim(0, 1)
  grfDA <- grf
  grf <- autoplot(ts(as.vector(dswminmax[nrow(dswminmax),]))) + theme_bw(base_size = 10) + geom_point(size=1)
  grf <- grf + theme(plot.title = element_blank())
  grf <- grf + theme(panel.grid.major = element_blank()) + theme(panel.grid.minor = element_blank())
  grf <- grf + ylab("")
  grf <- grf + xlab("sw min-max")
  grf <- grf + theme(plot.caption = element_text(hjust = 0.5))
  grf <- grf  + font
  grf <- grf + ylim(0, 1)
  grfMA <- grf
  grf <- grf + theme(plot.title = element_blank())
  grf <- grf + theme(panel.grid.major = element_blank()) + theme(panel.grid.minor = element_blank())
  grf <- grf + ylab("")
  grf <- grf + xlab("AN") + theme(axis.title.x = element_text(colour = "darkblue"))
  grf <- grf + theme(plot.caption = element_text(hjust = 0.5))
  grf <- grf  + font
  grf <- grf + ylim(0, 1)
  grfAA <- grf
  return(list(grfGA=grfGA, grfDA=grfDA, grfMA=grfMA, grfAA=grfAA))
}
ngminmax <- ts_norm_gminmax()
ngminmax <- fit(ngminmax, swy)
dgminmax <- transform(ngminmax, sw_range)
### ts_diff
ndiff <- ts_norm_diff()
ndiff <- fit(ndiff, swy)
ddiff <- transform(ndiff, sw_range)
### ts_swminmax()
nswminmax <- ts_norm_swminmax()
nswminmax <- fit(nswminmax, swy)
dswminmax <- transform(nswminmax, sw_range)
### ts_an()
n_an <- ts_norm_an()
n_an <- fit(n_an, swy)
d_an <- transform(n_an, sw_range)
list <- gen_graphics(dgminmax, ddiff, dswminmax, d_an)
list$grfGA <- list$grfGA + labs(caption = "(b)") + theme(plot.caption = element_text(colour = "darkgreen"))
list$grfDA <- list$grfDA + labs(caption = "(c)") + theme(plot.caption = element_text(colour = "darkgreen"))
list$grfMA <- list$grfMA + labs(caption = "(d)") + theme(plot.caption = element_text(colour = "darkgreen"))
list$grfAA <- list$grfAA + labs(caption = "(e)") + theme(plot.caption = element_text(colour = "darkgreen"))
```
### Visualization and Output
Plot the series with detected events and optionally save figures. Combine ggplot layers in one chunk for clarity.
```{r}
#mypng(file="figures/chap2_norm.png", width=1920, height=1080) 
gridExtra::grid.arrange(grfS, list$grfGA, list$grfDA, list$grfMA, list$grfAA,
                        layout_matrix = matrix(c(1,1,1,1,2,3,4,5), byrow = TRUE, ncol = 4))
#dev.off() 
```
## References
* General: Box, G. E. P. & Jenkins, G. (1970). Time Series Analysis.
* Ogasawara, E., Salles, R., Porto, F., Pacitti, E. Event Detection in Time Series. Springer, 2025. doi:10.1007/978-3-031-75941-3.
