---
title: "Chapter 2: Trend Diff"
output: html_document
---
```{r libs, message=FALSE, warning=FALSE}
library(ggplot2)
library(RColorBrewer)
library(tseries)
library(TSPred)
library(lmtest)
source("https://raw.githubusercontent.com/eogasawara/TSED/refs/heads/main/code/header.R")
```
## Theoretical Overview
This example focuses on trend removal via linear regression, moving-average smoothing, and differencing.
## Example Overview and Goals
We will: set up libraries, load data, remove trend with LR and MAS, apply first difference and PCT, and visualize results.
### What You Will Do
You will: compute residuals after LR and MAS detrending, apply differencing and percent-change transforms, and compare plots.
### Setup and Libraries
Short rationale for the libraries and any project-specific sources.
```{r setup, message=FALSE, warning=FALSE}
```
### Other Steps
Additional supporting steps that glue the workflow.
### Data Loading and Prep
Read the dataset and perform any minimal preparation required for modeling.
```{r data_loading}
data(examples_harbinger)
```
### Other Steps
Additional supporting steps that glue the workflow.
```{r detrend_lr}
data <- examples_harbinger$global_temperature_yearly
data$event <- FALSE
bp.test <- function(serie) {
  data <- data.frame(x = 1:length(serie), y = serie)
  fit <- lm(y ~ x, data = data)
  return(bptest(fit))
}
nonstationary.test <- function(serie) {
  return(data.frame(adf = round(adf.test(serie)$p.value, 2),
                    PP = round(PP.test(as.vector(serie))$p.value, 2),
                    bp = round(bp.test(serie)$p.value, 2)))
}
y <- ts(data$serie, start = c(1850, 1))
model <- lm(y ~ time(y))
```
### Visualization and Output
Plot the series with detected events and optionally save figures. Combine ggplot layers in one chunk for clarity.
```{r viz_lr}
grf <- autoplot(ts(model$residuals, start = c(1850, 1)))
```
### Other Steps
Additional supporting steps that glue the workflow.
```{r viz_lr_layers}
grf <- grf + theme_bw(base_size = 10)
grf <- grf + theme(plot.title = element_blank())
grf <- grf + theme(panel.grid.major = element_blank()) + theme(panel.grid.minor = element_blank())
grf <- grf + ylab("residual")
grf <- grf + xlab("time")
grf <- grf + labs(caption = "(a) LR trend removal") 
grf <- grf + theme(plot.caption = element_text(hjust = 0.5))
grf <- grf + geom_point(size = 0.25, col="black") 
grf <- grf  + font
grfa <- grf
y <- data$serie
y_transformed <- c(rep(NA, 9), TSPred::mas(y, 10))
```
### Visualization and Output
Plot the series with detected events and optionally save figures. Combine ggplot layers in one chunk for clarity.
```{r viz_mas_residual}
grf <- autoplot(ts(y - y_transformed, start = c(1850, 1)))
```
### Other Steps
Additional supporting steps that glue the workflow.
```{r viz_mas_layers}
grf <- grf + theme_bw(base_size = 10)
grf <- grf + theme(plot.title = element_blank())
grf <- grf + theme(panel.grid.major = element_blank()) + theme(panel.grid.minor = element_blank())
grf <- grf + ylab("residual")
grf <- grf + xlab("time")
grf <- grf + labs(caption = "(b) MAS trend removal") 
grf <- grf + theme(plot.caption = element_text(hjust = 0.5))
grf <- grf + geom_point(size = 0.25, col="black") 
grf <- grf  + font
grfb <- grf
y <- c(NA, diff(data$serie))
```
### Visualization and Output
Plot the series with detected events and optionally save figures. Combine ggplot layers in one chunk for clarity.
```{r viz_diff}
grf <- autoplot(ts(y, start = c(1850, 1)))
```
### Other Steps
Additional supporting steps that glue the workflow.
```{r viz_diff_layers}
grf <- grf + theme_bw(base_size = 10)
grf <- grf + theme(plot.title = element_blank())
grf <- grf + theme(panel.grid.major = element_blank()) + theme(panel.grid.minor = element_blank())
grf <- grf + ylab("transformation")
grf <- grf + xlab("time")
grf <- grf + labs(caption = "(c) First order differencing") 
grf <- grf + theme(plot.caption = element_text(hjust = 0.5))
grf <- grf + geom_point(size = 0.25, col="black") 
grf <- grf  + font
grfc <- grf
#PCT
pct_transformed <- TSPred::pct(data$serie)
yhat <- TSPred::pct.rev(pct_transformed,  attr(pct_transformed, "xi"))
nonstationary.test(pct_transformed)
```
### Visualization and Output
Plot the series with detected events and optionally save figures. Combine ggplot layers in one chunk for clarity.
```{r viz_pct}
grf <- autoplot(ts(pct_transformed, start = c(1850, 1)))
```
### Other Steps
Additional supporting steps that glue the workflow.
```{r viz_pct_layers}
grf <- grf + theme_bw(base_size = 10)
grf <- grf + theme(plot.title = element_blank())
grf <- grf + theme(panel.grid.major = element_blank()) + theme(panel.grid.minor = element_blank())
grf <- grf + ylab("transformation")
grf <- grf + xlab("time")
grf <- grf + labs(caption = "(d) PCT") 
grf <- grf + theme(plot.caption = element_text(hjust = 0.5))
grf <- grf + geom_point(size = 0.25, col="black") 
grf <- grf  + font
grfd <- grf
```
### Visualization and Output
Plot the series with detected events and optionally save figures. Combine ggplot layers in one chunk for clarity.
```{r}
#mypng(file="figures/chap2_trend_diff.png", width = 1600, height = 1080) #144 #720*1.5
gridExtra::grid.arrange(grfa, grfb, grfc, grfd,
                        layout_matrix = matrix(c(1,2,3,4), byrow = TRUE, ncol = 2))
#dev.off()  
```
## References
* General: Box, G. E. P. & Jenkins, G. (1970). Time Series Analysis.
* Ogasawara, E., Salles, R., Porto, F., Pacitti, E. Event Detection in Time Series. Springer, 2025. doi:10.1007/978-3-031-75941-3.
