---
title: "Chapter 1: Ygt Detect Chow"
output: html_document
---
```{r libs, message=FALSE, warning=FALSE}
library(ggplot2)
library(RColorBrewer)
library(daltoolbox)
library(harbinger)
source("https://raw.githubusercontent.com/eogasawara/TSED/refs/heads/main/code/header.R")
```
## Theoretical Overview
This example applies the Chow test for structural breaks. The Chow test assesses whether regression parameters differ before and after a candidate breakpoint, indicating a changepoint in the series.
## Example Overview and Goals
We walk through: setting up libraries, loading data, configuring and fitting the Chow detector, running detection, inspecting results, and visualizing the outcome.
### What You Will Do
You will: prepare the environment, load a dataset, configure the Chow-based detector, fit it, detect structural breaks, inspect detected events, and visualize the results.
### Setup and Libraries
Load project utilities and required packages.
```{r setup, message=FALSE, warning=FALSE}
# Shared helpers (themes, saving utilities, etc.)
```
### Data Loading and Prep
Read the dataset and prepare it for modeling.
```{r data_loading}
# Load example time series
data(examples_harbinger)
# Select the yearly global temperature series and initialize labels
data <- examples_harbinger$global_temperature_yearly
data$event <- FALSE
```
### Model Configuration
Define the Chow-based changepoint detector.
```{r model_config}
# Chow test for potential structural breaks
model <- hcp_chow()
```
### Fit the Model
Train (fit) the detector to the time series.
```{r model_fit}
model <- fit(model, data$serie)
```
### Event Detection
Run the detector to obtain candidate breakpoints.
```{r detect}
detection <- detect(model, data$serie)
```
### Inspect Results
Display detected events, if any, for a quick check.
```{r inspect}
print(detection |> dplyr::filter(event == TRUE))
```
### Visualization and Output
Plot the series with detected events and save the figure.
```{r visualize}
grf <- har_plot(model, data$serie, detection, data$event, idx = data$i) +
  font +
  scale_x_date(
    breaks = "10 years",
    date_labels = "%Y",
    limits = c(as.Date("1850-01-01"), as.Date("2030-01-01"))
  ) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
#save_png(grf, "figures/chap1_ygt_detect_chow.png", 1280, 720)
grf
```
## References
* Chow, G. C. (1960). Tests of equality between sets of coefficients in two linear regressions.
* Ogasawara, E., Salles, R., Porto, F., Pacitti, E. Event Detection in Time Series. Springer, 2025. doi:10.1007/978-3-031-75941-3.
