---
title: "Chapter 5: Discords via SAX"
output: html_document
---
```{r libs, message=FALSE, warning=FALSE}
library(ggplot2)
library(RColorBrewer)
library(daltoolbox)
library(harbinger)
source("https://raw.githubusercontent.com/eogasawara/TSED/refs/heads/main/code/header.R")
```
## Theoretical Overview
SAX (Symbolic Aggregate approXimation) converts time series to symbolic strings by discretizing PAA averages with normal-distribution breakpoints. Discord detection in the SAX domain identifies unusual subsequences whose symbolic patterns deviate from typical ones.
## Example Overview and Goals
We load an ECG-like example, configure a SAX-based discord detector, fit it, run detection, print the detected discord indices, and plot the results.
### Knitr Options
Keep output clean and reproducible.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```
### Setup and Libraries
Load project helpers and required packages.
```{r libs}
```
### Data
Load the motif/discord example series.
```{r data}
data(examples_motifs)
```
### Model, Fit, and Detect
Configure the SAX-based discord detector and run detection.
```{r model-detect}
data <- examples_motifs$mitdb102
rownames(data) <- 1:nrow(data)
data$event <- FALSE
# hdis_sax parameters (typical):
# - alphabet size (e.g., 26 symbols)
# - window size for subsequences (e.g., 25)
model <- hdis_sax(26, 25)
model <- fit(model, data$serie)
detection <- detect(model, data$serie)
print(detection[detection$event, ])
```
### Visualization and Output
Plot the series with detected discords and save the figure.
```{r plot}
grf <- har_plot(model, data$serie, detection) + font
#save_png(grf, "figures/chap5_discords_sax.png", 1280, 720)
grf
```
## References
* Yeh, C.-C. M., et al. (2016). Matrix Profile I: All Pairs Similarity Joins for Time Series.
* Ogasawara, E., Salles, R., Porto, F., Pacitti, E. Event Detection in Time Series. Springer, 2025. doi:10.1007/978-3-031-75941-3.
