---
title: "Chapter 4: KLDM (Kullback–Leibler Distance Monitor)"
output: html_document
---
```{r libs, message=FALSE, warning=FALSE}
library(ggplot2)
library(RColorBrewer)
library(daltoolbox)
library(harbinger)
library(heimdall)
source("https://raw.githubusercontent.com/eogasawara/TSED/refs/heads/main/code/header.R")
```
## Theoretical Overview
KLDM monitors distributional change by estimating the divergence (e.g., KL) between reference and recent windows of a stream, flagging drift when divergence exceeds a threshold.
## Example Overview and Goals
We feed the raw signal to KLDM in an online fashion, collect a detection frame, and visualize the flagged indices.
### Knitr Options
Keep output clean and reproducible.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```
### Setup and Libraries
Load helpers and detectors.
```{r libs}
```
### Data
Load the change-point example series.
```{r data}
data(examples_changepoints)
data <- examples_changepoints$complex
data$event <- NULL
```
### Online Detection
Run KLDM sequentially and collect detections.
```{r detect}
model <- dfr_kldist(target_feat = 'serie')
detection <- c()
state <- list(obj = model, pred = FALSE)
for (i in seq_along(data$serie)) {
  state <- update_state(state$obj, data$serie[i])
  if (state$drift) {
    type <- 'changepoint'
    state$obj <- reset_state(state$obj)
  } else {
    type <- ''
  }
  detection <- rbind(detection, list(idx = i, event = state$drift, type = type))
}
detection <- as.data.frame(detection)
```
### Visualization and Output
Plot detections over the series and save the figure.
```{r plot}
grf <- har_plot(model, data$serie, detection) + ylab("value")
#save_png(grf, "figures/chap4_kldm.png", 1280, 720)
grf
```
## References
* General: Box, G. E. P. & Jenkins, G. (1970). Time Series Analysis.
* Ogasawara, E., Salles, R., Porto, F., Pacitti, E. Event Detection in Time Series. Springer, 2025. doi:10.1007/978-3-031-75941-3.
