---
title: "Chapter 4: Multivariate PCA"
output: html_document
---
```{r libs, message=FALSE, warning=FALSE}
library(ggplot2)
library(RColorBrewer)
library(daltoolbox)
library(harbinger)
library(heimdall)
source("https://raw.githubusercontent.com/eogasawara/TSED/refs/heads/main/code/header.R")
```
## Theoretical Overview
Principal Component Analysis (PCA) projects multivariate data into lower-dimensional components that capture the most variance. Change points may become more evident in a univariate projection (e.g., PC1).
## Example Overview and Goals
We demonstrate a complete, reproducible workflow: setting up libraries, loading data, configuring a detector, fitting it, running detection, optionally evaluating, and visualizing the results.
### Knitr Options
Keep output clean and reproducible.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```
### Setup and Libraries
Load helpers and required packages.
```{r libs}
```
### Data Generation
Create a simple 2D dataset and a boolean drift region based on quadrant consistency.
```{r data-gen}
set.seed(1)
n <- 500
data <- data.frame(serie1 = rnorm(n),
                   serie2 = rnorm(n))
m1 <- mean(data$serie1)
m2 <- mean(data$serie2)
data$drift <- ((data$serie1 > m1) & (data$serie2 > m2)) |
              ((data$serie1 < m1) & (data$serie2 < m2))
```
### Visualization and Output
Plot the series with detected events and optionally save figures. Combine ggplot layers in one chunk for clarity.
```{r plot-scatter}
grf <- ggplot(data, aes(x = serie1, y = serie2)) +
  geom_rect(xmin = m1, xmax = +Inf, ymin = m2, ymax = +Inf, fill = "lightgray", alpha = 0.2) +
  geom_rect(xmin = -Inf, xmax = m1, ymin = -Inf, ymax = m2, fill = "lightgray", alpha = 0.2) +
  geom_point(size = 2) +
  theme_classic()
plot(grf)
```
### Train/Test Framing
Order non-drift first, then drift; add index and one ground-truth event.
```{r prep-series}
ts_before <- data[data$drift == FALSE, ]
ts_after  <- data[data$drift == TRUE, ]
data <- rbind(ts_before, ts_after)
data$i <- seq_len(nrow(data))
data$event <- FALSE
data$event[min(which(data$drift))] <- TRUE
drift <- which(data$event)
model <- fit(hcp_chow(), data$serie1)
```
### Event Detection (Series 1)
```{r detect-serie1}
detection <- detect(model, data$serie1)
print(detection$idx[detection$event])
```
### Visualization and Output
Plot the series with detected events and optionally save figures. Combine ggplot layers in one chunk for clarity.
```{r plot-serie1}
grfA <- ggplot(data, aes(x = i, y = serie1)) +
  geom_line() +
  theme_classic()
```
### Event Detection (Series 2)
```{r detect-serie2}
model <- fit(hcp_chow(), data$serie2)
detection <- detect(model, data$serie2)
print(detection$idx[detection$event])
```
### Visualization and Output
Plot the series with detected events and optionally save figures. Combine ggplot layers in one chunk for clarity.
```{r plot-serie2}
grfB <- ggplot(data, aes(x = i, y = serie2)) +
  geom_line() +
  theme_classic()
```
### PCA Projection
Compute PC1 from standardized features and fit Chow test on the projection.
```{r pca-proj}
scaled_data <- base::scale(data[, 1:2])       # standardize features
pca_result  <- stats::princomp(scaled_data)   # PCA
pca <- as.data.frame(pca_result$scores)$'Comp.1'
model <- fit(hcp_chow(), pca)
```
### Event Detection (PCA)
```{r detect-pca}
detection <- detect(model, pca)
print(detection$idx[detection$event])
```
### Visualization and Output
Plot the series with detected events and optionally save figures. Combine ggplot layers in one chunk for clarity.
```{r plot-pca}
grfPCA <- ggplot(data, aes(x = i, y = pca)) +
  geom_line() +
  geom_vline(xintercept = drift, linetype = "dotted", color = "black", linewidth = 1) +
  theme_classic()
```
### Save Figure
```{r save-figure}
#mypng(file = "figures/chap4_multivariate_pca.png", width = 1280, height = 1080)
gridExtra::grid.arrange(grfA, grfB, grfPCA,
                        layout_matrix = matrix(c(1, 2, 3), byrow = TRUE, ncol = 1))
#dev.off()
```
## References
* General: Box, G. E. P. & Jenkins, G. (1970). Time Series Analysis.
* Ogasawara, E., Salles, R., Porto, F., Pacitti, E. Event Detection in Time Series. Springer, 2025. doi:10.1007/978-3-031-75941-3.
