---
title: "Chapter 4: Drift Type"
output: html_document
---
```{r libs, message=FALSE, warning=FALSE}
library(ggplot2)
library(RColorBrewer)
library(daltoolbox)
library(harbinger)
source("https://raw.githubusercontent.com/eogasawara/TSED/refs/heads/main/code/header.R")
```
## Theoretical Overview
We contrast different concept-drift patterns: sudden (abrupt), gradual, recurring, and periodic perturbations. Understanding these types helps match detectors to the expected drift dynamics.
## Example Overview and Goals
We synthesize four short series exhibiting different drift types and annotate segments to highlight the change patterns.
### Knitr Options
Make output clean and reproducible.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```
### Setup and Libraries
Load helpers and packages.
```{r libs}
```
### Data and Generators
```{r data-gen}
set.seed(1)
data_a <- function() {
  data <- NULL
  data <- c(data, c(2, 1, 2, 3, 2, 1, 2, 3)) 
  data <- c(data, c(2, 1, 2, 3, 2, 1, 2, 3)) 
  data <- c(data, 10+c(2, 1, 2, 3, 2, 1, 2, 3))
  data <- c(data, 10+c(2, 1, 2, 3, 2, 1, 2, 3))
  return(data+10)
}
data_b <- function() {
  data <- NULL
  data <- c(data, c(2, 1, 2, 3, 2, 1, 2, 3)) 
  data <- c(data, c(0, 0, 0, 0, 1.2, 2.4, 3.6, 4.8)+c(2, 1, 2, 3, 2, 1, 2, 3)) 
  data <- c(data, c(6.0,7.2,8.4,9.6,10,10,10,10)+c(2, 1, 2, 3, 2, 1, 2, 3))
  data <- c(data, 10+c(2, 1, 2, 3, 2, 1, 2, 3))
  return(data+10)
}
data_c <- function() {
  data <- NULL
  data <- c(data, c(2, 1, 2, 3, 2, 1, 2, 3)) 
  data <- c(data, c(0, 0, 10, 0, 0, 10, 0, 0)+c(2, 1, 2, 3, 2, 1, 2, 3)) 
  data <- c(data, c(10, 10, 0, 10, 10, 0, 10, 10)+c(2, 1, 2, 3, 2, 1, 2, 3))
  data <- c(data, 10+c(2, 1, 2, 3, 2, 1, 2, 3))
  return(data+10)
}
data_d <- function() {
  data <- NULL
  data <- c(data, c(2, 1, 2, 3, 2, 1, 2, 3)) 
  data <- c(data, 10+c(2, 1, 2, 3, 2, 1, 2, 3))
  data <- c(data, c(2, 1, 2, 3, 2, 1, 2, 3)) 
  data <- c(data, 10+c(2, 1, 2, 3, 2, 1, 2, 3))
  return(data+10)
}
a <- data_a()
model <- fit(harbinger(), a)
detection <- detect(model, a)
```
```{r annotate-a}
detection$event[16] <- TRUE
detection$type[16]  <- "anomaly"
```
```{r plot-a}
grf <- har_plot(model, a, detection)
```
```{r decorate-a}
grf <- grf + scale_x_continuous(breaks=seq(0,32,8))
grf <- grf + ylab(" ") + ylim(0, 30)
grf <- grf + geom_segment(aes(x = 16, y = 8, xend = 16, yend = 25), col = "black", linetype = "dashed")
grf <- grf + font 
grf <- grf + ggplot2::annotate(geom="text", x=16, y=1, label="(a)", color="black")
grf <- grf + ggplot2::annotate(geom="text", x=4, y=29, label="SD[1]", color="darkblue", parse=TRUE)
grf <- grf + ggplot2::annotate(geom="text", x=12, y=29, label="SD[2]", color="darkblue", parse=TRUE)
grf <- grf + ggplot2::annotate(geom="text", x=20, y=29, label="SD[3]", color="darkblue", parse=TRUE)
grf <- grf + ggplot2::annotate(geom="text", x=28, y=29, label="SD[4]", color="darkblue", parse=TRUE)
grf <- grf + geom_segment(aes(x = 0, y = 27, xend = 0, yend = 30), col = "darkblue")
grf <- grf + geom_segment(aes(x = 8, y = 27, xend = 8, yend = 30), col = "darkblue")
grf <- grf + geom_segment(aes(x = 16, y = 27, xend = 16, yend = 30), col = "darkblue")
grf <- grf + geom_segment(aes(x = 24, y = 27, xend = 24, yend = 30), col = "darkblue")
grf <- grf + geom_segment(aes(x = 32, y = 27, xend = 32, yend = 30), col = "darkblue")
grfA <- grf
b <- data_b()
model <- fit(harbinger(), b)
detection <- detect(model, b)
```
```{r annotate-b}
detection$event[c(10,20)] <- TRUE
detection$type[c(10,20)]  <- "anomaly"
```
```{r plot-b}
grf <- har_plot(model, b, detection)
```
```{r decorate-b}
grf <- grf + scale_x_continuous(breaks=seq(0,32,8))
grf <- grf + ylab(" ") + ylim(0, 30)
grf <- grf + geom_segment(aes(x = 8, y = 8, xend = 8, yend = 25), col = "black", linetype = "dashed")
grf <- grf + geom_segment(aes(x = 24, y = 8, xend = 24, yend = 25), col = "black", linetype = "dashed")
grf <- grf + font
grf <- grf + ggplot2::annotate(geom="text", x=16, y=1, label="(b)", color="black")
grf <- grf + ggplot2::annotate(geom="text", x=4, y=29, label="SD[1]", color="darkblue", parse=TRUE)
grf <- grf + ggplot2::annotate(geom="text", x=12, y=29, label="SD[2]", color="darkblue", parse=TRUE)
grf <- grf + ggplot2::annotate(geom="text", x=20, y=29, label="SD[3]", color="darkblue", parse=TRUE)
grf <- grf + ggplot2::annotate(geom="text", x=28, y=29, label="SD[4]", color="darkblue", parse=TRUE)
grf <- grf + geom_segment(aes(x = 0, y = 27, xend = 0, yend = 30), col = "darkblue")
grf <- grf + geom_segment(aes(x = 8, y = 27, xend = 8, yend = 30), col = "darkblue")
grf <- grf + geom_segment(aes(x = 16, y = 27, xend = 16, yend = 30), col = "darkblue")
grf <- grf + geom_segment(aes(x = 24, y = 27, xend = 24, yend = 30), col = "darkblue")
grf <- grf + geom_segment(aes(x = 32, y = 27, xend = 32, yend = 30), col = "darkblue")
grfB <- grf
c <- data_c()
model <- fit(harbinger(), c)
detection <- detect(model, c)
```
```{r annotate-c}
detection$event[c(10,23)] <- TRUE
detection$type[c(10,23)]  <- "anomaly"
```
```{r plot-c}
grf <- har_plot(model, c, detection)
```
```{r decorate-c}
grf <- grf + scale_x_continuous(breaks=seq(0,32,8))
grf <- grf + ylab(" ") + ylim(0, 30)
grf <- grf + geom_segment(aes(x = 8, y = 8, xend = 8, yend = 25), col = "black", linetype = "dashed")
grf <- grf + geom_segment(aes(x = 24, y = 8, xend = 24, yend = 25), col = "black", linetype = "dashed")
grf <- grf + font
grf <- grf + ggplot2::annotate(geom="text", x=16, y=1, label="(c)", color="black")
grf <- grf + ggplot2::annotate(geom="text", x=4, y=29, label="SD[1]", color="darkblue", parse=TRUE)
grf <- grf + ggplot2::annotate(geom="text", x=12, y=29, label="SD[2]", color="darkblue", parse=TRUE)
grf <- grf + ggplot2::annotate(geom="text", x=20, y=29, label="SD[3]", color="darkblue", parse=TRUE)
grf <- grf + ggplot2::annotate(geom="text", x=28, y=29, label="SD[4]", color="darkblue", parse=TRUE)
grf <- grf + geom_segment(aes(x = 0, y = 27, xend = 0, yend = 30), col = "darkblue")
grf <- grf + geom_segment(aes(x = 8, y = 27, xend = 8, yend = 30), col = "darkblue")
grf <- grf + geom_segment(aes(x = 16, y = 27, xend = 16, yend = 30), col = "darkblue")
grf <- grf + geom_segment(aes(x = 24, y = 27, xend = 24, yend = 30), col = "darkblue")
grf <- grf + geom_segment(aes(x = 32, y = 27, xend = 32, yend = 30), col = "darkblue")
grfC <- grf
d <- data_d()
model <- fit(harbinger(), d)
detection <- detect(model, d)
```
```{r annotate-d}
detection$event[c(8,16,24)] <- TRUE
detection$type[c(8,16,24)]  <- "anomaly"
```
```{r plot-d}
grf <- har_plot(model, d, detection)
```
### Other Steps
Additional supporting steps that glue the workflow.
```{r}
grf <- grf + scale_x_continuous(breaks=seq(0,32,8))
grf <- grf + ylab(" ") + ylim(0, 30)
grf <- grf + geom_segment(aes(x = 8, y = 8, xend = 8, yend = 25), col = "black", linetype = "dashed")
grf <- grf + geom_segment(aes(x = 16, y = 8, xend = 16, yend = 25), col = "black", linetype = "dashed")
grf <- grf + geom_segment(aes(x = 24, y = 8, xend = 24, yend = 25), col = "black", linetype = "dashed")
grf <- grf + font
grf <- grf + ggplot2::annotate(geom="text", x=16, y=1, label="(d)", color="black")
grf <- grf + ggplot2::annotate(geom="text", x=4, y=29, label="SD[1]", color="darkblue", parse=TRUE)
grf <- grf + ggplot2::annotate(geom="text", x=12, y=29, label="SD[2]", color="darkblue", parse=TRUE)
grf <- grf + ggplot2::annotate(geom="text", x=20, y=29, label="SD[3]", color="darkblue", parse=TRUE)
grf <- grf + ggplot2::annotate(geom="text", x=28, y=29, label="SD[4]", color="darkblue", parse=TRUE)
grf <- grf + geom_segment(aes(x = 0, y = 27, xend = 0, yend = 30), col = "darkblue")
grf <- grf + geom_segment(aes(x = 8, y = 27, xend = 8, yend = 30), col = "darkblue")
grf <- grf + geom_segment(aes(x = 16, y = 27, xend = 16, yend = 30), col = "darkblue")
grf <- grf + geom_segment(aes(x = 24, y = 27, xend = 24, yend = 30), col = "darkblue")
grf <- grf + geom_segment(aes(x = 32, y = 27, xend = 32, yend = 30), col = "darkblue")
grfD <- grf
```
```{r save-figure}
#mypng(file = "figures/chap4_drift_type.png", width = 1280, height = 720)
gridExtra::grid.arrange(grfA, grfB, grfC, grfD,
                        layout_matrix = matrix(c(1,1,2,2,3,3,4,4), byrow = TRUE, ncol = 4))
#dev.off()
```
## References
* General: Box, G. E. P. & Jenkins, G. (1970). Time Series Analysis.
* Ogasawara, E., Salles, R., Porto, F., Pacitti, E. Event Detection in Time Series. Springer, 2025. doi:10.1007/978-3-031-75941-3.
